\chapter{Introduction}

This book is one volume in a series consisting of photographs I have taken, selected from a lifetime of taking pictures. 

I have never been a professional, but I was and am committed to the art of photography. I am a ``street photographer'' open to abstraction. I shoot what catches my eye, and I prize beauty. When I take pictures of people, I usually prefer they don't notice. 

I take pictures so I can see more clearly what God has created.

These pictures were variously taken with 35 mm film cameras, a variety of digital cameras, and smartphones. They were taken in my native state of Utah, other states including Washington, California, and New York, and countries around the world. I have tended to use the sharpest possible camera that is small enough to carry with me at all times. For some years now I have been using the Sony RX100 series. Increasingly, however, I am finding that my smartphone is a real camera.

The images are all in natural color. They are all used as far as possible without image manipulation; in a few cases, I have leveled horizons or removed spots. I shoot in JPEG rather than RAW. Selected metadata from the photographs is printed. If the photograph is a digital scan of a 35 mm slide, the creation date is the date of that scan. Captions are incomplete and cryptic. Images are, to the best of my recollection where timestamps are lacking, in chronological order.

\emph{All pictures are included in the full resolution with which they were taken}. Thus, you can zoom into any image to see more detail. Pictures copied out of this book will also be in full resolution. As the file size of this book would otherwise become unmanageable thanks to the use of uncompressed images, it has been split into volumes of no more than about 500 megabytes each. Even so, the files are huge. 

Although pictures in portrait format are not rotated to be larger on the page, they nevertheless also have the full resolution with which they were taken and can be zoomed into or printed at full resolution. You can probably rotate images in portrait format on your screen to get a bigger image for viewing.

Metadata from the images, where readable and relevant, is printed beneath the caption. If by some chance the reader wants to know more about the technicalities, they can look up the meaning and units of the tag values from the EXIF tag name and number, using some resource such as \href{https://www.cipa.jp/std/documents/e/DC-008-2012_E.pdf}{\emph{\textbf{Exchangeable image file format for digital still cameras: Exif Version 2.3}}}.

I invite the reader to copy pictures out of this book for printing and viewing, but not for commercial use or re-publication. 

\begin{table}[h]
\centering
\captionsetup{labelformat=empty}
\caption{\textbf{Maximum Recommended Print Sizes}}
\begin{tabular}{lrr}
\hline
\emph{Source of image}                                  & \emph{Close up} & \emph{On a wall} \\
\hline
35 mm Velvia slide (for comparison) & 16" x 20" & 18" x 24" \\
My scans of slides                            & 11" x 14"  & 16" x 20"  \\
20 megapixel digital camera             & 16" x 20"  & 18" x 24"  \\
High-end smartphone                      & 8" x 10"   & Up to 16" x 20"  \\
\hline
\end{tabular}
\end{table}

As I take more pictures that I think are good enough to be in this series, I will add new volumes.
