\documentclass[english,11pt,letterpaper,onecolumn]{scrartcl}

%\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{csquotes}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathptmx}
\usepackage{enumitem}
% Extra leading.
\renewcommand{\baselinestretch}{1.125}
\usepackage{tocloft}
% \usepackage{fancyhdr}
\usepackage{scrlayer-scrpage}
\usepackage{ifthen}
\usepackage{keyval}
\usepackage{geometry}
\usepackage{url}
\usepackage{calc}
\usepackage{array}
\usepackage{graphicx}
\usepackage{color}
\usepackage{listings}
\usepackage{supertabular}
% definitions used by included articles, reproduced here for
% educational benefit, and to minimize alterations needed to be made
% in developing this sample file.
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amscd}     % commutative diagrams
\usepackage{mathrsfs}  % This package allows the use of script letters. Type \mathscr to invoke math script.
\usepackage{amsbsy}
%\usepackage{scrpage2}
\usepackage[pdftex,
colorlinks=true,
linkcolor=blue,
pdfpagelabels,
pdfstartpage=3
]{hyperref}
% \usepackage{poemscol}
% \global\verselinenumbersfalse
\makeindex
\definecolor{LstColor}{cmyk}{0.1,0.1,0,0.025}
\setcounter{tocdepth}{9}
\newcommand\floor[1]{\lfloor#1\rfloor}
\newcommand\ceil[1]{\lceil#1\rceil}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\hypersetup{breaklinks=true,
bookmarks=true,
pdfauthor={},
pdftitle={},
colorlinks=true,
citecolor=blue,
urlcolor=blue,
linkcolor=magenta,
pdfborder={0 0 0}}

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

\usepackage{version}%
\setcounter{MaxMatrixCols}{30}

%%%% Packages added - Peter
\usepackage{color}
\usepackage{amscd}     % commutative diagrams
\usepackage{mathrsfs}  % This package allows the use of script letters.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newtheorem{theorem}{Theorem}[section]
%\theoremstyle{plain}
\newtheorem{acknowledgement}{Acknowledgement}
\newtheorem{algorithm}{Algorithm}[section]
\newtheorem{axiom}{Axiom}[section]
\newtheorem{case}{Case}[section]
\newtheorem{claim}{Claim}[section]
\newtheorem{conclusion}{Conclusion}[section]
\newtheorem{condition}{Condition}[section]
\newtheorem{conjecture}{Conjecture}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{criterion}{Criterion}[section]
\newtheorem{definition}{Definition}[section]
\newtheorem{example}{Example}[section]
\newtheorem{exercise}{Exercise}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{notation}{Notation}[section]
\newtheorem{problem}{Problem}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{remark}{Remark}[section]
\newtheorem{solution}{Solution}[section]
\newtheorem{summary}{Summary}[section]
\numberwithin{equation}{section}
\excludeversion{comment1}

%%%%% Macros added - Peter
\newcommand{\st}{\,|\,}
\newcommand{\C}{\mathbb{C}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\I}{\mathbb{I}}
\newcommand{\bH}{\mathbb{H}}
\newcommand{\K}{\mathbb{K}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\X}{\mathbb{X}}
\newcommand{\Y}{\mathbb{Y}}
\newcommand{\Z}{\mathbb{Z}}
%
\newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}}
\newcommand{\mcD}{\mathcal{D}}
\newcommand{\cE}{\mathcal{E}}
\newcommand{\cF}{\mathcal{F}}
\newcommand{\cG}{\mathcal{G}}
\newcommand{\calH}{\mathcal{H}}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\cK}{\mathcal{K}}
\newcommand{\mcL}{\mathcal{L}}
\newcommand{\cM}{\mathcal{M}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}}
\newcommand{\calR}{\mathcal{R}}
\newcommand{\cS}{\mathcal{S}}
\newcommand{\mcT}{\mathcal{T}}
\newcommand{\cU}{\mathcal{U}}
\newcommand{\cW}{\mathcal{W}}
\newcommand{\cY}{\mathcal{Y}}
\newcommand{\bcF}{\boldsymbol{\mathcal{F}}}
%
\newcommand{\oY}{\overline{Y}}
\newcommand{\uY}{\underline{Y}}
%
\renewcommand{\Re}{\mathop{\mathrm{Re}}}
\renewcommand{\Im}{\mathop{\mathrm{Im}}}
%
\newcommand{\card}{\mathop{\mathrm{card}}}
\newcommand{\Int}{\mathop{\mathrm{int}}}
\newcommand{\eps}{\varepsilon}
\newcommand{\kap}{\varkappa}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\inn}[2]{{\langle #1,#2 \rangle}}
\newcommand{\wt}[1]{{\widetilde{#1}}}

\newcommand{\conv}{\mathrm{conv}\,}
\newcommand{\bx}{{\boldsymbol{x}}}
\newcommand{\by}{{\boldsymbol{y}}}
\newcommand{\bk}{{\boldsymbol{k}}}
\newcommand{\bm}{{\boldsymbol{m}}}
\newcommand{\bc}{{\boldsymbol{c}}}
\newcommand{\ba}{{\boldsymbol{a}}}
\newcommand{\bp}{{\boldsymbol{p}}}
\newcommand{\bP}{{\mathbb{P}}}
\newcommand{\bS}{{\boldsymbol{S}}}
\newcommand{\bq}{{\boldsymbol{q}}}
\newcommand{\bfe}{{\boldsymbol{e}}}
\newcommand{\bone}{{\boldsymbol{1}}}
\newcommand{\bu}{{\boldsymbol{u}}}
\newcommand{\bv}{{\boldsymbol{v}}}
\newcommand{\bw}{{\boldsymbol{w}}}
\newcommand{\bff}{{\boldsymbol{f}}}
\newcommand{\bkk}{{\boldsymbol{k-1}}}
\newcommand{\bxi}{{\boldsymbol{Xi}}}
\newcommand{\balpha}{{\boldsymbol{\alpha}}}
\newcommand{\bbeta}{{\boldsymbol{\beta}}}
\newcommand{\bgamma}{{\boldsymbol{\gamma}}}
\newcommand{\blambda}{{\boldsymbol{\lambda}}}
\newcommand{\bmu}{{\boldsymbol{\mu}}}
\newcommand{\gr}{{\mathrm{graph\,}}}
\newcommand{\of}{\overline{f}}
\newcommand{\og}{\overline{g}}
\newcommand{\oc}{\overline{c}}
\newcommand{\ou}{\overline{u}}
\newcommand{\ox}{\overline{x}}
\newcommand{\oX}{\overline{X}}
\newcommand{\oXi}{\overline{X}_i}
\newcommand{\uc}{\underline{c}}
\newcommand{\oy}{\overline{y}}
\newcommand{\uy}{\underline{y}}
\newcommand{\odelta}{\overline{\delta}}
\newcommand{\udelta}{\underline{\delta}}
\newcommand{\oPhi}{\overline{\Phi}}
\newcommand{\map}{\textrm{Map}\,}
\newcommand{\loc}{\mathrm{loc}}
\newcommand{\mydot}{\;\cdot\;}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\range}{range}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\Aff}{Aff}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\diag}{diag}

\usepackage[
backend=biber,
bibencoding=utf8,
style=numeric,
sorting=ynt,
hyperref=true,
backref=true
]{biblatex}
\addbibresource{gogins.bib}
\begin{document}

\title{Self-Consciousness a Fixed Point}
\author{Michael Gogins \\ \texttt{michael.gogins@gmail.com}}
\maketitle
%\pagestyle{scrheadings}

\begin{abstract}
We first identify a family of philosophical arguments in which
self-consciousness is explicitly or implicitly treated as a fixed point. What
this means will be developed in detail. We then demonstrate that consciousness
as a fixed point is not consistent with naturalism or mechanism.
\end{abstract}

%\lohead{Parametric Composition}



\section{Introduction}

\lstset{language=c++,basicstyle=\ttfamily\tiny,commentstyle=\ttfamily\tiny,tabsize=2,breaklines,backgroundcolor=\color{LstColor},fontadjust=true,keepspaces=false,showstringspaces=false,moredelim=[is][\textbf]{\\emph\{}{\}}}.

In ordinary language, to say that I am self-conscious is to say that I am
conscious \textit{that} I am conscious. There are a number of philososphical
arguments, some historically important, some famous, some wrong, and some not
refuted, that use this basic notion of self-consciousness as a starting point.

To say that self-consciousness is a fixed point is to say that to be conscious
that I am conscious does not involve recursion or an infinite regress.
Phenomenologically, there is no difference between being self-conscious, and
being conscious of being self-conscious. We will consider why this appearance
might or might not reflect reality.

Please note, to say that self-consciousness is a fixed point is not the same as
the \textit{cogito}. But if self-consciousness is in reality a fixed point, that
has a number of interesting implications, some of them quite important. Among
them are that in self-consciousness there is no distinction between subject and
object, there is in it no passage of time, and self-consciousness is of an
irreducible simplicity, similar to or perhaps even identical with the divine
simplicity ascribed to God.

The main reason for affirming that self-consciousness is in reality a fixed
point is the absurdity of affirming otherwise. If this appearance is a delusion
or epiphenomenon, then how does one become conscious of this delusion? But there
are other reasons to affirm it. Self-consciousness is adaptive. Whatever other
reasons self-consciousness might have for existing, it undoubtedly is, or
reflects, a biological adaptation that confers a selective advantage. This is
evident not only among human beings, but, if the mirror test is a reliable
indicator of self-consciousness, also among other higher animals, each at or
near the top of the food chain in its environment, including other great apes,
some cetaceans, and some birds.

The adaptive function of self-consciousness is apparent, at least to some
extent, in what organisms that lack it cannot do. They cannot use symbolic
reasoning, they cannot solve abstract problems, and they cannot form voluntary
social relationships. The facts of biology demonstrate, beyond any reasonable
doubt, that self-consciousness confers, together with the ability to make and
use tools, an overwhelming selective advantage.



\printbibliography
\end{document}
